---
title: "Online Resource 01 - Supplementary appendix E"
author: "John Dagsvik, Mariachiara Fortuna, Sigmund H. Moen"
subtitle: "How does the temperature vary over time? Evidence on the Stationary and Fractal nature of Temperature Fluctuations"
output:
  pdf_document: default

---


\vspace{8.5cm}

**Affiliations:** 

John K. Dagsvik, Statistics Norway, Research Department; 

Mariachiara Fortuna, freelance statistician, Turin; 

Sigmund Hov Moen, Westerdals Oslo School of Arts, Communication and Technology. 


\vspace{0.5cm}

**Corresponding author:**

John K. Dagsvik, E-mail: john.dagsvik@ssb.no

Mariachiara Fortuna, E-mail: mariachiara.fortuna1@gmail.com (reference for code and analysis)


\newpage


```{r global options, echo=F}
knitr::opts_chunk$set(
  echo = F, # Set to TRUE for pdf with code
  message=FALSE,
  warning = FALSE)
longrun <- FALSE
```



```{r loading libraries, warning = F, message = F}
require(tempFGN)
require(tidyverse)
require(tidyr)
require(lubridate)
library(maps)
```


```{r path building, hide=T}
# DATA PATH
data_final_path <- file.path("data","final")
data_supporting_path <- file.path("data", "supporting")
data_moberg_path <- file.path("data", "moberg")

# OUTPUT PATH
output_supporting_path <- file.path("output", "supporting")
output_table_path <- file.path("output", "table")
output_figure_path <- file.path("output", "figure")
output_temporary_path <- file.path("output", "temporary")
output_manipuated_path <- file.path("output", "manipulated")
```


```{r data reading}

# ACCESS TO ALL THE TIME SERIES
all <- read.csv(file.path(data_supporting_path, "T0.TempInfo.csv"), sep=";", dec=",")
country_all <- all$Country
station_all <- all$Station
nja <- nrow(all)
data_dir_all <- file.path(data_final_path, country_all, paste0(station_all, ".txt"))
stationame_all <- paste0(country_all,", ",station_all)

```
  



```{r FE1 - Final appendix sheet, fig.height=9}

for (j in 1:nja){ 
  
  data <- read.delim(data_dir_all[j], header=F, na.strings=99)
  dataM <- monthlyAdj(data, scale=T)
  Zm <- dataM$Zm 
  Ym <- cumsum(Zm)
  
  par(oma=c(0, 0, 2, 0))
  
  par(fig=c(0, 0.5, 0.66, 0.96), cex=0.6)
  maps::map('world', ylim=c(-60,90), mar=c(2,2,2,2))
  points(all$Lon[j], all$Lat[j], col=2, pch=16, cex=2)
  
  par(fig=c(0.5, 1, 0.62, 0.98), cex=0.6, 
      bty="l", mar=c(6,4,5,2), new=T)
  autocorrPlot(Zm, H="cfH", 
               cex.axis=1, cex.lab=1, cex.main=1.1, cex.dots=1.2)$plot
  
  par(fig=c(0, 1, 0.33, 0.63), mar=c(3,4,5,2), bty="l", cex=0.7, new=T)
  blueRedPlot(Time=dataM$Time, Zj= Zm, break.val=1960, ic=NULL)
  
  par(fig=c(0,0.5,0,0.33), cex=0.7, new=T)
  fgtSelfSim(Yj=Ym)$plot
  
  par(fig=c(0.5,1,0,0.33), cex=0.7, new=T)
  fgtNormality(Yj=Ym)$plot
  
  title(stationame_all[j], outer=T, cex.main=1.5)
}

```



